Operational Feature Selection with Gaussian Mixture Models for remote sensing images

### 2016-10-06 paper.tex
update

        A sub-GMM model can be estimated at a reduced complexity when one fold is remove from the training set. From~\cite{fauvel2015fast}, the following formulae for unbiased estimators can be derived: %% TODO: rajouter des \hat{} partout: ce sont des estimateurs.
        \begin{prop}
            \label{eq:update-cv1}
            (Cross-validation mean update)
            \begin{equation*}
                \boldsymbol{\mu}_c^{n_c-\nu_c} = \frac{n_c \boldsymbol{\mu}_c^{n_c} - \nu_c \boldsymbol{\mu}_c^{\nu_c}}{n_c - \nu_c} \nonumber
            \end{equation*}
        \end{prop}
        \begin{prop}
            \label{eq:update-cv2}
            (Cross-validation covariance matrix update)
            \begin{align*}
                \boldsymbol{\Sigma}_c^{n_c-\nu_c} = &\frac{1}{n_c-\nu_c-1} \biggl( (n_c-1) \boldsymbol{\Sigma}_c^{n_c} - (\nu_c-1) \boldsymbol{\Sigma}_c^{\nu_c} \nonumber \\
                &- \frac{n_c \nu_c}{(n_c-\nu_c)} (\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})(\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})^t \biggr) \nonumber
            \end{align*}
        \end{prop}
        \noindent where $n_c$ is the number of samples of class $c$, $\nu_c$ is the number of samples of class $c$ removed from the initial set, exposants on $\boldsymbol{\Sigma}_c$ and $\boldsymbol{\nu}_c$ denotes the set of samples used to compute them.

@@

        \begin{equation}\label{eq:cov}
            \boldsymbol{\Sigma}^{(k)} =
            \bigg[\begin{array}{cc}
            \boldsymbol{\Sigma}^{(k-1)} & \mathbf{u}      \\
            \mathbf{u}^t          & \sigma_{kk} \\
            \end{array}\bigg],
        \end{equation}
        where  $\mathbf{u}$  is  the  $k^{th}$ column  of  the  matrix
        without      the      diagonal      element,      \emph{i.e.},
        $\mathbf{u}_{i}   =    \boldsymbol{\Sigma}^{(k)}_{i,k}$   with
        $i  \in [1,k-1]$.  Using  block matrix  inverse formulae,  the
        following proposition holds:
        \begin{prop}
        \label{eq:update-inv}
            (Forward update rule for inverse of covariance matrix)
            \begin{equation}\label{eq:cov:inv}
                (\boldsymbol{\Sigma}^{(k)})^{-1} =
                \bigg[\begin{array}{cc}
                \mathbf{A} & \mathbf{v} \\
                \mathbf{v}^t  & \frac{1}{\alpha} \\
                \end{array}\bigg]
            \end{equation}
        \end{prop}
        \noindent                                                where
        $\mathbf{A}     =     (\boldsymbol{\Sigma}^{(k-1)})^{-1}     +
        \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}
        \mathbf{u}^t              (\boldsymbol{\Sigma}^{(k-1)})^{-1}$,
        $\mathbf{v}           =           -           \frac{1}{\alpha}
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}       \mathbf{u}$       and
        $      \alpha      =      \sigma_{kk}      -      \mathbf{u}^t
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}  \mathbf{u}$.  This  update
        formulae      is     used      to      obtain     all      the
        $(\boldsymbol{\Sigma}^{(k)})^{-1}$  corresponding  to all  the
        possible augmented set of a given selection iteration. Similar
        update formulae can be written for the backward step.

        % In the case of a backward step in SFFS algorithm, the formula is inverted in order to compute $(\boldsymbol{\Sigma}^{(k-1)})^{-1}$ knowing $(\boldsymbol{\Sigma}^{(k)})^{-1}$. The update rule becomes (proof in Appendix~\ref{app:proof-update})
        % \begin{prop}
        % \label{eq:update-inv-back}
        %     (Backward update rule for inverse of covariance matrix)
        %     \begin{equation*}
        %         (\boldsymbol{\Sigma}^{(k-1)})^{-1} = \underbrace{\mathbf{A}}_{\substack{\text{computed once}\\ \text{per selection step}}} - \underbrace{\alpha \mathbf{v} \mathbf{v}^t}_{\substack{\text{computed for} \\ \text{each augmented set}}}
        %     \end{equation*}
        % \end{prop}

        Using eq.~\ref{eq:cov} and~\ref{eq:cov:inv}, it is possible to
        deduce  the   following  propositions  (proof  are   given  in
        Appendix~\ref{app:proof-update}).
        \begin{prop}
        \label{eq:update-quad}
            (Update rule for quadratical term)
            \begin{align}
                (\mathbf{x}^{(k)})^t (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)} = &\underbrace{(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)}}_{\substack{\text{computed once}\\ \text{per selection step}}} \nonumber \\
                &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{align}
        \end{prop}
        \begin{prop}
        \label{eq:update-log}
            (Update rule for logdet)
            \begin{equation}
                \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) = \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)}_{\substack{\text{computed once}\\ \text{per selection step}}} + \underbrace{\log \alpha}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{equation}
        \end{prop}

@@

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 50,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu-big}}
        \begin{tabular}{|lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-7}
            GMM SFS kappa & 0.713 (0.001) & 0.684 (0.001) & 0.531 (0.005) & 20000 & 340 & 29 \\
            GMM SFS JM &    0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6 & 330 & 10 \\
            GMM SFFS JM &   0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6.6 & 340 & 10 \\
            GMM ridge &     0.641 (0.015) & 0.611 (0.026) & 0.440 (0.015) & 460 & 2000 & all \\
            Random Forest & {\bfseries 0.851 (0.001)} & {\bfseries 0.715 (0.001)} & {\bfseries 0.573 (0.002)} & 2000 & 2000 & all \\
            \bottomrule
        \end{tabular}
    \end{table}

@@

\begin{proof}[Proof of proposition~\ref{eq:update-quad}]
  \begin{align*}
    (&\mathbf{x}^{(k)})^t (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)}
       = \left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t   & x_k \end{array}\right]
                                                            \left[\begin{array}{cc}
                                                                    \mathbf{A}   & \mathbf{v} \\
                                                                    \mathbf{v}^t & \frac{1}{\alpha}
                                                                  \end{array}\right]
                                                                                   \left[\begin{array}{c} \mathbf{v} \\ x_k \end{array}\right] \\
     &= \left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t \mathbf{A} + x_k \mathbf{v}^t & (\mathbf{x}^{(k-1)})^t \mathbf{v} + \frac{x_k}{\alpha} \end{array}\right]
                                                                                       \left[\begin{array}{c} \mathbf{v} \\ x_k \end{array}\right] \\
     &= (\mathbf{x}^{(k-1)})^t \mathbf{A} \mathbf{x}^{(k-1)} + x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + (\mathbf{x}^{(k-1)})^t \mathbf{v} x_k + \frac{(x_k)^2}{\alpha} \\
     &= (\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}^{(k-1)})^{-1} + \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u} \mathbf{u}^t (\boldsymbol{\Sigma}^{(k-1)})^{-1}) \mathbf{x}^{(k-1)}\\
     &~~~+ 2 x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha} \\
     &= (\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}^{(k-1)})^{-1} + \alpha \mathbf{v} \mathbf{v}^t) \mathbf{x}^{(k-1)}\\
     &~~~+ 2 x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha} \\
     &= (\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( (\mathbf{x}^{(k-1)})^t \mathbf{v} \mathbf{v}^t \mathbf{x}^{(k-1)} \\
     &~~~+ 2 \frac{x_k}{\alpha} \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha^2}) \\
     &= (\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( (\mathbf{x}^{(k-1)})^t \mathbf{v} + \frac{x_k}{\alpha} )^2 \\
     &= (\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2
  \end{align*}
\end{proof}

>>>>

        A sub-GMM model can be estimated at a reduced complexity when one fold is remove from the training set. From~\cite{fauvel2015fast}, the following formulae for unbiased estimators can be derived:
        \begin{prop}[Cross-validation mean update]
            \label{eq:update-cv1}
            \begin{equation*}
                \boldsymbol{\hat{\mu}}_c^{n_c-\nu_c} = \frac{n_c \boldsymbol{\hat{\mu}}_c^{n_c} - \nu_c \boldsymbol{\hat{\mu}}_c^{\nu_c}}{n_c - \nu_c} \nonumber
            \end{equation*}
        \end{prop}
        \begin{prop}[Cross-validation covariance matrix update]
            \label{eq:update-cv2}
            \begin{align*}
                \boldsymbol{\hat{\Sigma}}_c^{n_c-\nu_c} = &\frac{1}{n_c-\nu_c-1} \biggl( (n_c-1) \boldsymbol{\hat{\Sigma}}_c^{n_c} - (\nu_c-1) \boldsymbol{\hat{\Sigma}}_c^{\nu_c} \nonumber \\
                &- \frac{n_c \nu_c}{(n_c-\nu_c)} (\boldsymbol{\hat{\mu}}_c^{\nu_c}-\boldsymbol{\hat{\mu}}_c^{n_c})(\boldsymbol{\hat{\mu}}_c^{\nu_c}-\boldsymbol{\hat{\mu}}_c^{n_c})^t \biggr) \nonumber
            \end{align*}
        \end{prop}
        \noindent where $n_c$ is the number of samples of class $c$, $\nu_c$ is the number of samples of class $c$ removed from the initial set, exposants on $\boldsymbol{\Sigma}_c$ and $\boldsymbol{\nu}_c$ denotes the set of samples used to compute them.

@@

        \begin{equation}\label{eq:cov}
            \boldsymbol{\Sigma}^{(k)} =
            \bigg[\begin{array}{cc}
            \boldsymbol{\Sigma}^{(k-1)} & \mathbf{u}      \\
            \mathbf{u}^t          & \sigma_{kk} \\
            \end{array}\bigg],
        \end{equation}
        where  $\mathbf{u}$  is  the  $k^{th}$ column  of  the  matrix
        without      the      diagonal      element,      \emph{i.e.},
        $\mathbf{u}_{i}   =    \boldsymbol{\Sigma}^{(k)}_{i,k}$   with
        $i  \in [1,k-1]$.  Using  block matrix  inverse formulae,  the
        following proposition holds:
        \begin{prop}[Forward update rule for inverse of covariance matrix]
        \label{eq:update-inv}
            \begin{equation}\label{eq:cov:inv}
                (\boldsymbol{\Sigma}^{(k)})^{-1} =
                \bigg[\begin{array}{cc}
                \mathbf{A} & \mathbf{v} \\
                \mathbf{v}^t  & \frac{1}{\alpha} \\
                \end{array}\bigg]
            \end{equation}
        \end{prop}
        \noindent                                                where
        $\mathbf{A}     =     (\boldsymbol{\Sigma}^{(k-1)})^{-1}     +
        \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}
        \mathbf{u}^t              (\boldsymbol{\Sigma}^{(k-1)})^{-1}$,
        $\mathbf{v}           =           -           \frac{1}{\alpha}
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}       \mathbf{u}$       and
        $      \alpha      =      \sigma_{kk}      -      \mathbf{u}^t
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}  \mathbf{u}$.  This  update
        formulae      is     used      to      obtain     all      the
        $(\boldsymbol{\Sigma}^{(k)})^{-1}$  corresponding  to all  the
        possible augmented set of a given selection iteration. Similar
        update formulae can be written for the backward step.

        % In the case of a backward step in SFFS algorithm, the formula is inverted in order to compute $(\boldsymbol{\Sigma}^{(k-1)})^{-1}$ knowing $(\boldsymbol{\Sigma}^{(k)})^{-1}$. The update rule becomes (proof in Appendix~\ref{app:proof-update})
        % \begin{prop}[Backward update rule for inverse of covariance matrix]
        % \label{eq:update-inv-back}
        %     \begin{equation*}
        %         (\boldsymbol{\Sigma}^{(k-1)})^{-1} = \underbrace{\mathbf{A}}_{\substack{\text{computed once}\\ \text{per selection step}}} - \underbrace{\alpha \mathbf{v} \mathbf{v}^t}_{\substack{\text{computed for} \\ \text{each augmented set}}}
        %     \end{equation*}
        % \end{prop}

        Using eq.~\ref{eq:cov} and~\ref{eq:cov:inv}, it is possible to
        deduce  the   following  propositions  (proof  are   given  in
        Appendix~\ref{app:proof-update}).
        \begin{prop}[Update rule for quadratical term]
        \label{eq:update-quad}
            \begin{align}
                (\mathbf{x}^{(k)})^t (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)} = &\underbrace{(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)}}_{\substack{\text{computed once}\\ \text{per selection step}}} \nonumber \\
                &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{align}
        \end{prop}
        \begin{prop}[Update rule for logdet]
        \label{eq:update-log}
            \begin{equation}
                \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) = \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)}_{\substack{\text{computed once}\\ \text{per selection step}}} + \underbrace{\log \alpha}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{equation}
        \end{prop}

@@

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 50,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu-big}}
        \begin{tabular}{lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-7}
            GMM SFS kappa & 0.713 (0.001) & 0.684 (0.001) & 0.531 (0.005) & 20000 & 340 & 29 \\
            GMM SFS JM &    0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6 & 330 & 10 \\
            GMM SFFS JM &   0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6.6 & 340 & 10 \\
            GMM ridge &     0.641 (0.015) & 0.611 (0.026) & 0.440 (0.015) & 460 & 2000 & all \\
            Random Forest & {\bfseries 0.851 (0.001)} & {\bfseries 0.715 (0.001)} & {\bfseries 0.573 (0.002)} & 2000 & 2000 & all \\
            \bottomrule
        \end{tabular}
    \end{table}

@@

\begin{proof}[Proof of proposition~\ref{eq:update-quad}]
    \begin{alignat*}{3}
    (&\mathbf{x}^{(k)})^t && (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)}
       = \left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t   & x_k \end{array}\right]
                                                            \left[\begin{array}{cc}
                                                                    \mathbf{A}   & \mathbf{v} \\
                                                                    \mathbf{v}^t & \frac{1}{\alpha}
                                                                  \end{array}\right]
                                                                                   \left[\begin{array}{c} \mathbf{v} \\ x_k \end{array}\right] \\
     &= &&\left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t \mathbf{A} + x_k \mathbf{v}^t & (\mathbf{x}^{(k-1)})^t \mathbf{v} + \frac{x_k}{\alpha} \end{array}\right]
                                                                                       \left[\begin{array}{c} \mathbf{v} \\ x_k \end{array}\right] \\
     &= &&(\mathbf{x}^{(k-1)})^t \mathbf{A} \mathbf{x}^{(k-1)} + x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + (\mathbf{x}^{(k-1)})^t \mathbf{v} x_k + \frac{(x_k)^2}{\alpha} \\
     &= &&(\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}^{(k-1)})^{-1} + \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u} \mathbf{u}^t (\boldsymbol{\Sigma}^{(k-1)})^{-1}) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha} \\
     &= &&(\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}^{(k-1)})^{-1} + \alpha \mathbf{v} \mathbf{v}^t) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha} \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( (\mathbf{x}^{(k-1)})^t \mathbf{v} \mathbf{v}^t \mathbf{x}^{(k-1)} \\
     & &&+ 2 \frac{x_k}{\alpha} \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha^2}) \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( (\mathbf{x}^{(k-1)})^t \mathbf{v} + \frac{x_k}{\alpha} )^2 \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2
    \end{alignat*}
\end{proof}

\begin{proof}[Proof of proposition~\ref{eq:update-log}]
    \begin{align*}
        \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) =
        &\log \left(\left|
            \bigg[\begin{array}{cc}
                \boldsymbol{\Sigma}^{(k-1)} & \mathbf{u}      \\
                \mathbf{u}^t          & \sigma_{kk} \\
            \end{array}\bigg]
        \right|\right) \\
        &\log \left(\left|
            \bigg[\begin{array}{cc}
                \boldsymbol{\Sigma}^{(k-1)} & \mathbf{0}_{k-1,1}     \\
                \mathbf{u}^t          & 1 \\
            \end{array}\bigg]
            \bigg[\begin{array}{cc}
                \mathbf{I}_{k-1} & (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}      \\
                \mathbf{0}_{k-1,1}^t          & \sigma_{kk} - \mathbf{u}^t(\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}\\
            \end{array}\bigg]
        \right|\right) \\
        &\log \left(\left|
            \bigg[\begin{array}{cc}
                \boldsymbol{\Sigma}^{(k-1)} & \mathbf{0}_{k-1,1}     \\
                \mathbf{u}^t          & 1 \\
            \end{array}\bigg]
        \right|\right)
        +\log \left(\left|
            \bigg[\begin{array}{cc}
                \mathbf{I}_{k-1} & (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}      \\
                \mathbf{0}_{k-1,1}^t          & \alpha \\
            \end{array}\bigg]
        \right|\right) \\
        &\log \left( |\boldsymbol{\Sigma}^{(k-1)}| |1| \right) + \log \left( |\mathbf{I}_{k-1}| |\alpha| \right)\\
        &\log \left( |\boldsymbol{\Sigma}^{(k-1)}| \right) + \log ( \alpha )
    \end{align*}
\end{proof}
